### 2025-09-18 basic-statistics.qmd
20250918 - ANOVA formula

>>>>

The *F* statistic (which is used to determine statistical significance) in ANOVA is computed by first partitioning the sum of squares as in @eq-partitionSumSquares:

$$
SS_{\text{total}} = SS_{\text{between}} + SS_{\text{within}}
$$ {#eq-partitionSumSquares}

where $SS_{\text{total}}$ is the total sum of squares, which is equal to @eq-sumSquaresTotal:

$$
SS_{\text{total}} = \sum_{i = 1}^{k} \sum_{j = 1}^{n_i} \left( X_{ij} - \bar{X}_{..} \right)^2
$$ {#eq-sumSquaresTotal}

where $X_{ij}$ is the raw data point (i.e., observation $j$ within group $i$), $n_i$ is the sample size of group $i$, $\bar{X}_{i.}$ is the mean of group $i$, and $\bar{X}_{..}$ is the grand mean (the mean across all observations from all groups).

$SS_{\text{between}}$ is the between-groups sum of squares, which is equal to @eq-sumSquaresBetween:

$$
SS_{\text{between}} = \sum_{i = 1}^{k} n_i \left( \bar{X}_{i.} - \bar{X}_{..} \right)^2
$$ {#eq-sumSquaresBetween}

and $SS_{\text{within}}$ is the within-groups sum of squares, which is equal to @eq-sumSquaresWithin:

$$
SS_{\text{within}} = \sum_{i = 1}^{k} \sum_{j = 1}^{n_i} \left( X_{ij} - \bar{X}_{i.} \right)^2
$$ {#eq-sumSquaresWithin}

Mean squares are then computed from the sum of squares as in @eq-meanSquares:

$$
\begin{aligned}
  MS_{\text{between}} &= \frac{SS_{\text{between}}}{k - 1} \\
  MS_{\text{within}} &= \frac{SS_{\text{within}}}{N - k}
\end{aligned}
$$ {#eq-meanSquares}

where $k$ is the number of groups and $N$ is the sample size.

Then, the *F* statistic is computed as the ratio of between-groups mean squares to within-groups mean squares as in @eq-fStatistic:

$$
F = \frac{MS_{\text{between}}}{MS_{\text{within}}}
$$ {#eq-fStatistic}

In sum, ANOVA examines the ratio of between-groups sum of squares to within-groups sum of squares (after adjusting for degrees of freedom).
